---
title: "Lecture-06 Examples"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook covers basic map production in `R` using a variety of tools for producing *static* maps (as opposed to the interactive maps `leaflet` makes).

## Dependencies
This notebook requires a variety of packages for working with spatial data:

```{r load-packages}
# tidyverse packages
library(ggplot2)      # plotting data

# spatial packages
library(mapview)      # preview spatial data
library(tmap)         # map layouts
library(sf)           # spatial data tools

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
library(viridis)      # color palettes
```

## Getting Help
There are a number of ways to get help in `R`. We'll go over:

* accessing help files with `?st_read`
* searching for help
* posting questions on the discussion forum

## Analysis Development - README files
Each assignment from here on out should include a `README` file that documents the directories and files within a project. This can be created by going to `File > New File > Text File` in RStudio. Save your new document in the top-level of your project with the name `README.md`. Be sure to include the `.md` file extension - RStudio will not automatically add it to your file.

Your `README` file should include a couple of things. First, include a sentance that describes the overall purpose of your project:

> The goal of the `lab-05` project is to create static maps of structures at risk from severe weather in the St. Louis metropolitan region.

Then, include an outline of the key components of your project. Document what is in the `data/`, `docs/`, and `results/` subdirectories in your project. It is best to do this as an itemized list in Markdown:

```markdown
* `data/` - both shapefiles needed for the lab
* `docs/` - the lab notebook
* `results/` - the two maps required as final outputs
```

## Load Data (Lecture-01 Review)
This notebook requires the data stored in `data/example-data/`. Remember that we use `sf::st_read()` to load shapefile data:

```{r load-data}
city <- st_read(here("data", "example-data", "STL_BOUNDARY_City", "STL_BOUNDARY_City.shp"), stringsAsFactors = FALSE)
nhoods <- st_read(here("data", "example-data", "STL_DEMOS_Nhoods", "STL_DEMOS_Nhoods.shp"), stringsAsFactors = FALSE)
```

## Projections
We briefly reviewed this last week - we need to ensure our data our projected correctly (but will get into the weeds on this at a later date). To ensure that our data are projected correctly, we use `sf::st_transform()` to project both using the UTM 15N projected coordinate system:

```{r project-data}
# city boundary
city <- st_transform(city, crs = 32615)

# neighborhood demographics
nhoods <- st_transform(nhoods, crs = 32615)
```

## Simple Maps with `ggplot2`
### Basic Mapping of Geometric Objects
`ggplot2` is the premier graphics package for `R`. It is an incredibly powerful visualization tool that increasingly supports spatial work and mapping. The basic `ggplot2` workflow requires chaining together functions with the `+` sign. 

We'll start by creating a `ggplot2` object with the `ggplot()` function, and then adding a "geom", which provides `ggplot2` instructions on how our data should be visualized. We can read these like paragraphs:

1. First, we create an empty `ggplot2` object, **then**
2. we add the `nhoods` data and visualize its geometry.

```{r ggplot2-nhoodSimple}
ggplot() +
  geom_sf(data = nhoods, fill = "#bababa")
```

You can see empty spaces where there are major parks - if we wanted to give these a background color, we could add the `city` layer under our `nhoods` layer. We can also add the `city` layer again on top to give the city border a pronounced outline. `ggplot2` relies on layering different geoms to produce complicated plots. We can assign each geom a specific set of aesthetic characteristics and use data from different objects.

```{r ggplot2-nhoodSimple2}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = NA) +
  geom_sf(data = nhoods, fill = "#bababa") +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75)
```

### Mapping Quantities with `ggplot2`
If we wanted to start to map data instead of just the geometric properties, we would specify an "aesthetic mapping" using `mapping= aes()` in the geom of interest. Here, we create a fill that is the product of taking the population in 2017 and normalizing it by square kilometers as we did in the `leaflet` section above. We provide additional instructions about how our data should be colored with the `scale_fill_distiller()` function, which gives us access to the `RColorBrewer` palettes.

```{r ggplot2-nhood1}
# create ggplot object
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = NA) +
  geom_sf(data = nhoods, mapping = aes(fill = pop17/(AREA/1000000))) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_distiller(palette = "Greens", trans = "reverse") -> ggplot_17_1

# print object
ggplot_17_1
```

This map also stores our `ggplot` object in its own space in our global environment. This allows us the ability to update it later, and to more easily save it.

### Creating Map Layouts with `ggplot2`

Before we save it, however, we should create a more substantial layout. We'll use the `name` argument in `scale_fill_distiller()` to name the legend, the `labs()` function to add text to our layout, and `theme_minimal()` to remove some of the default `ggplot2` theme elements:

```{r ggplot2-nhood2}
# create ggplot object
ggplot() +
  geom_sf(data = city, fill = "#ededed", color = NA) +
  geom_sf(data = nhoods, mapping = aes(fill = pop17/(AREA/1000000))) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_distiller(palette = "Greens", trans = "reverse", name = "Population per\nSquare Kilometer") +
  labs(
    title = "Population Density (2017)",
    subtitle = "Neighborhoods in St. Louis, MO",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() -> ggplot_17_2

# print object
ggplot_17_2
```

Next, to save our map, we use the `ggsave()` function:

```{r ggplot2-save-nhood2}
ggsave(here("examples", "results", "ggplot2_popDensity17.png"), ggplot_17_2, dpi = 500)
```

We can repeat this process for the 1950 data:

```{r ggplot2-nhood3}
# create ggplot object
ggplot() +
  geom_sf(data = city, fill = "#ededed", color = NA) +
  geom_sf(data = nhoods, mapping = aes(fill = pop50/(AREA/1000000))) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_distiller(palette = "Blues", trans = "reverse", name = "Population per\nSquare Kilometer") +
  labs(
    title = "Population Density (1950)",
    subtitle = "Neighborhoods in St. Louis, MO",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() -> ggplot_17_3

# print object
ggplot_17_3
```

To save our map, we again use the `ggsave()` function:

```{r ggplot2-save-nhood3}
ggsave(here("examples", "results", "ggplot2_popDensity50.png"), ggplot_17_3, dpi = 500)
```

### Using `viridis` with `ggplot2`

The other option for color palettes is the `viridis` family of palettes. These are specified by replacing `scale_fill_distiller()` with `scale_fill_viridis()`. The `option` argument replaces `palette`, but `name` has the same functionality:

```{r ggplot2-nhood4}
# create ggplot object
ggplot() +
  geom_sf(data = city, fill = "#ededed", color = NA) +
  geom_sf(data = nhoods, mapping = aes(fill = pop17/(AREA/1000000))) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_viridis(option = "cividis", name = "Population per\nSquare Kilometer") +
  labs(
    title = "Population Density (2017)",
    subtitle = "Neighborhoods in St. Louis, MO",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() -> ggplot_17_4

# print object
ggplot_17_4
```

The other options for `viridis` are `viridis`, `magma`, `plasma`, and `inferno`:

```{r ggplot2-nhood5}
# create ggplot object
ggplot() +
  geom_sf(data = city, fill = "#ededed", color = NA) +
  geom_sf(data = nhoods, mapping = aes(fill = pop17/(AREA/1000000))) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_viridis(option = "viridis", name = "Population per\nSquare Kilometer") +
  labs(
    title = "Population Density (2017)",
    subtitle = "Neighborhoods in St. Louis, MO",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() -> ggplot_17_5

# print object
ggplot_17_5
```

To save our map, we again use the `ggsave()` function:

```{r ggplot2-save-nhood5}
ggsave(here("examples", "results", "ggplot2_popDensity17_2.png"), ggplot_17_5, dpi = 500)
```

## Managing Our Enviornment
With GIS work, our environment gets cluttered. As I work on an analysis, I find it useful to remove objects once I know that I am done with them. We use the `base::rm()` function to do this:

```{r clean-up-ggplot2-objects}
rm(ggplot_17_1, ggplot_17_2, ggplot_17_3, ggplot_17_4, ggplot_17_5)
```

## Map Layouts with `tmap`
`tmap` uses a similar logic to `ggplot2` - it layers elements on top of each other to produce maps. It is dedicated to working with spatial data, however, and has some features that `ggplot2` does not. 

### Basic Mapping of Geometric Objects
We'll start with a basic map that, like we have previously, just display the geometry of the city's neighborhoods. Similar to `ggplot2`, functions are chained together with the `+` sign. We can read these like paragraphs:

1. First, we take the `nhoods` data, **then**
2. we create our `tmap` layer out of its shape, **then**
3. we add a fill using our layer, **then**
4. we add borders using our layer.
  
```{r tmap-simple}
nhoods %>%
  tm_shape() +
    tm_fill() +
    tm_borders() 
```

### Mapping Quantities with `tmap`
Like `ggplot2`, we can plot quantities using the `tm_polygons()` function. The `palette` argument accepts names of both `RColorBrewer` and `viridis` palettes.

```{r tmap-quantities}
nhoods %>%
  tm_shape() +
    tm_polygons(col = "pop17", palette = "Greens")
```

Notice that this is a map of population counts, and is therefore not normalized. `tamp` makes the normalization process easy, with the `convert2density` argument:

```{r tmap-density}
nhoods %>%
  tm_shape() +
    tm_polygons(col = "pop17", palette = "Reds", convert2density = TRUE)
```

We can shrink (or grow) the number of classes using the `n` argument in `tm_polygons`, though I've found it to be unreliable occasionally:

```{r tmap-density2}
nhoods %>%
  tm_shape() +
    tm_polygons(col = "pop17", 
                palette = "BuPu", 
                n = 3,
                convert2density = TRUE)
```

We can also change the breaks are calculated. `tmap` uses the `"pretty"` approach by default, whereas ArcGIS uses the `"jenks"` approach. We can mirror ArcGIS by specifying `"jenks"`, and can continue to adjust the number of breaks: 

```{r tmap-jenks}
nhoods %>%
  tm_shape() +
    tm_polygons(col = "pop17", 
                palette = "BuPu", 
                style = "jenks",
                n = 6,
                convert2density = TRUE)
```

Another option is to use the `"equal"` approach to breaks, which divides our observations into equally sized classes:

```{r tmap-equal}
nhoods %>%
  tm_shape() +
    tm_polygons(col = "pop17", 
                palette = "BuPu", 
                style = "equal",
                n = 6,
                convert2density = TRUE)
```

### Creating Map Layouts with `tmap`
Once we have a map we like, we can begin to build a layout around it. Like with our `ggplot2` map layout, we'll add the city underneath by adding a shape below `nhoods`. We'll use the `city` data for this. We'll add the `city` on top as well to achieve that outline effect we discussed with `ggplot2` as well:

```{r tmap-add-background}
tm_shape(city) +
  tm_fill(fill = "#ebebeb") + 
  tm_shape(nhoods) +
  tm_polygons(col = "pop17", 
              palette = "viridis", 
              style = "jenks",
              convert2density = TRUE) +
  tm_shape(city) +
  tm_borders(lwd = 2)
```

Notice how we have to add each layer using `tm_shape()` before beginning to modify its atheistic properties.

We can also add adornments to our map layouts, including a scale bar (with `tm_scale_bar()`):

```{r tmap-add-scale-bar}
tm_shape(city) +
  tm_fill(fill = "#ebebeb") + 
  tm_shape(nhoods) +
  tm_polygons(col = "pop17", 
              palette = "viridis", 
              style = "jenks",
              convert2density = TRUE) +
  tm_shape(city) +
  tm_borders(lwd = 2) +
  tm_scale_bar() 
```

Once we have a layout that we like, we can use `tm_layout()` to add a title and move the legend if necessary.  

```{r tmap-layout1}
# create tmap object
tm_shape(city) +
  tm_fill(fill = "#ebebeb") + 
  tm_shape(nhoods) +
  tm_polygons(col = "pop17", 
              palette = "viridis", 
              style = "jenks",
              convert2density = TRUE,
              title = "Population per\nSquare Kilometer") +
  tm_shape(city) +
  tm_borders(lwd = 2) +
  tm_scale_bar() +
  tm_layout(
    title = "Population Density (2017)",
    frame = FALSE,
    legend.outside = TRUE,
    legend.position = c("left", "bottom")) -> tmap_17_1

# print object
tmap_17_1
```

`tmap` lacks the ability to add subtitles and captions to plot layouts, which is a drawback. Once we have our object created, we can save it using `tmap_save()`, which is functionally the same as `ggplot2::ggsave()` but with slightly different arguments:

```{r tmap-save-layout1}
tmap_save(tm = tmap_17_1, filename = here("examples", "results", "tamp_popDensity17_1.png"), dpi = 500)
```

### Adding Histograms
One neat feature that `tmap` has is the ability to add a histogram of the mapped variable to the legend as well. This is done by adding `legend.hist = TRUE` to the `tm_polygons()` function:

```{r tmap-layout2}
# create tmap object
tm_shape(city) +
  tm_fill(fill = "#ebebeb") + 
  tm_shape(nhoods) +
  tm_polygons(col = "pop17", 
              palette = "GnBu", 
              style = "jenks",
              convert2density = TRUE,
              title = "Population per\nSquare Kilometer",
              legend.hist = TRUE) +
  tm_shape(city) +
  tm_borders(lwd = 2) +
  tm_scale_bar() +
  tm_layout(
    title = "Population Density (2017)",
    frame = FALSE,
    legend.outside = TRUE,
    legend.position = c("left", "bottom")) -> tmap_17_2

# print object
tmap_17_2
```

Once again, we can save this using `tmap_save()`:

```{r tmap-save-layout2}
tmap_save(tm = tmap_17_2, filename = here("examples", "results", "tamp_popDensity17_2.png"), dpi = 500)
```



```{r move-to-docs, include=FALSE}
# you do need to include this in any notebook you create for this class
fs::file_copy(here::here("examples", "lecture-06.nb.html"), 
              here::here("docs", "index.nb.html"), 
              overwrite = TRUE)
```
