---
title: "Lecture-06 Examples"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook covers basic map production in `R` using a variety of tools for producing *static* maps (as opposed to the interactive maps `leaflet` makes).

## Dependencies
This notebook requires a variety of packages for working with spatial data:

```{r load-packages}
# tidyverse packages
library(ggplot2)      # plotting data
library(magrittr)     # pipe operator

# spatial packages
library(mapview)      # preview spatial data
library(tmap)         # map layouts
library(sf)           # spatial data tools

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
library(viridis)      # color palettes
```

## Load Data and Re-project
This notebook requires the data stored in `data/example-data/`. Remember that we use `sf::st_read()` to load shapefile data. This time, however, we'll combine our `st_read()` call with `st_transform()` in a pipeline. We'll need to load the pipe operator using the `magrittr` package. We use `magrittr` and not `dplyr` here because we do not need the extra features of `dplyr`. 

```{r load-data}
# city boundary
st_read(here("data", "example-data", "STL_BOUNDARY_City", 
             "STL_BOUNDARY_City.shp"), stringsAsFactors = FALSE) %>%
  st_transform(crs = 32615) -> city

# fire stations
st_read(here("data", "example-data", "STL_PUBLICSAFTEY_FireStations", 
             "STL_PUBLICSAFTEY_FireStations.shp"), stringsAsFactors = FALSE) %>%
  st_transform(crs = 32615) -> fire
```

## Exploring the Data
We can use `mapview()` to explore our point data:

```{r mapview}
mapview(fire)
```

We have two variables we'll be focusing on, the `stationID` variable (to simulate mapping point data with a quantitative attribute) and the `battalion`, which we'll use for creating our facets.

## Mapping Points
### Using `ggplot2`
There are two differences we need to be conscious of when we plot points using `ggplot2`. The first is that we need to use a different argument for applying colors and aesthetic mappings. Instead of using `fill` as we did with our polygon features, we want to use `color`. 

```{r ggplot2-stationsSimple}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, color = "#8b0000") 
```

With polygon data like the city boundary, the `color` arguments are used to control the border. With point data, the `color` argument is used to fill in the symbols.

#### Basic Options
We can extend this functionality by changing shapes. The default for `geom_sf()` is to use [shape 16](http://www.sthda.com/english/wiki/ggplot2-point-shapes). We can change shapes by using the `shape` argument:

```{r ggplot2-stationsShape}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, color = "#8b0000", shape = 12)
```

One technique we use in cartography is to add a black border around points to make them stand out. We can achieve this with shape 21 (for circular symbols):

```{r ggplot2-stationsFill}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, fill = "#8b0000", shape = 21) 
```

Notice here that we've switched back to using `fill`. If we also supplied a `color` argument, it would change the border color (as it does with polygon data). Finally, with point data, we can change the size of the symbol using the `size` argument:

```{r ggplot2-stationsSize}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, fill = "#8b0000", shape = 21, size = 2) 
```

#### Applying Aesthetic Mappings
We can also use asthetic mappings with these data, much as we did with polygon data. If we use any of the simple shapes, we use the `color` argument inside of our aesthetic mapping and either `scale_color_distiller()` or `scale_color_viridis()`:

```{r ggplot2-stationsAesthetic}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, mapping = aes(color = stationID), size = 2) +
  scale_color_distiller(palette = "Reds", trans = "reverse")
```

If we were to use the point symbol with a border, we would use the `fill` arguments instead along with either `scale_fill_distiller()` or `scale_fill_viridis()`:

```{r ggplot2-stationsAestheticFill}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, mapping = aes(fill = stationID), shape = 21, size = 2) +
  scale_fill_viridis(option = "inferno")
```

### Using `tmap`
We can achieve the same effects using the `tmap` package. A basic point map would use `tm_shape()` twice, once to add the city background and once to add the fire stations. We use `tm_polygons()` to customize the apperance of the city layer and `tm_symbols()` to customize the appearence of the points. 

```{r tmap-base}
tm_shape(city) +
  tm_polygons(col = "#ffffff", lwd = .75) +
  tm_shape(fire) +
  tm_symbols(col = "#8b0000", size = .5)
```

Similarly, we can use `tm_lines()` to customize line elements.

If we wanted to mirrior the aesthetic mappings from `ggplot2`, we can adjust our `tmap` call slightly. Instead of declaring a color in `tm_symbols(col)`, we specify a variable:

```{r tmap-final}
tm_shape(city) +
  tm_polygons(col = "#ffffff", lwd = .75) +
  tm_shape(fire) +
  tm_symbols(col = "stationID", palette = "Reds", size = .5)
```

## Small Multiples
We can map these in groups according to any characteristic (typically a categorical or ordinal variable) using a technique called faceting. Facets implement the idea of "small multiples" - multiple plots of the same scale (or spatial area) showing data according to the specified characteristics. We can do this in `ggplot2` using the `facet_grid()` and `facet_wrap()` functions. For what it is worth, `facet_wrap()` often producing more pleasing output:

```{r ggplot-facets}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = "#000000", size = .75) +
  geom_sf(data = fire, mapping = aes(fill = stationID), shape = 21, size = 2) +
  scale_fill_distiller(palette = "Reds", trans = "reverse") +
  labs(
    title = "Stations by Battalion",
    subtitle = "St. Louis Fire Department"
  ) +
  theme_minimal() +
  facet_wrap(~battalion)
```

Nothing should change about your call execept for the addition of `facet_wrap()` at the end of your call. Note the addition of the tilde (`~`) in the function call - thsi is important!

We can achieve a similar effect with the `tm_facets()` function in `tmap`. Instead of a tilde, we specify the grouping variable with the `by = ` argument:

```{r tm-facets}
tm_shape(city) +
  tm_polygons(col = "#ffffff", lwd = .75) +
  tm_shape(fire) +
  tm_symbols(col = "stationID", palette = "Reds", size = .5) +
  tm_facets(by = "battalion", free.coords = FALSE)
```

The `free.cords` argument ensures that our data are mapped to the same extent in each multiple. Without it, `tmap` will zoom in to the closest possible extent in each panel, which defeats the visual effect of faceting.

```{r move-to-docs, include=FALSE}
# you do need to include this in any notebook you create for this class
fs::file_copy(here::here("examples", "lecture-06.nb.html"), 
              here::here("docs", "additional-examples.nb.html"), 
              overwrite = TRUE)
```
