---
title: "Lab-05-Replication"
author: "Branson Fox"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook replicates the results of Lab-05.

## Dependencies
This notebook requires the following packages to load our data and create maps

```{r load-packages}
# tidyverse packages
library(ggplot2)      # data visualization

# spatial packages
library(sf)           # tools for spatial data
library(mapview)      # interactive mapping preview
library(tmap)         # thematic map making

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
library(viridis)      # color palettes
```

## Load Data
This notebook requires data from the Lecture-06 repository, that we should have put in the `data/` directory of our project.

```{r load-data}
counties <- st_read(here("data", "METRO_BOUNDARY_Counties", "METRO_BOUNDARY_Counties.shp"), stringsAsFactors = FALSE)
mobileHome <- st_read(here("data", "METRO_STRUCTURE_PctMobileHome", "METRO_STRUCTURE_PctMobileHome.shp"), stringsAsFactors = FALSE)
```

## Part 1: Data Exploration
### Question 1
We will start by exploring the variables in our mobileHome data using the `str()` function.

```{r p1-q1}
str(mobileHome)
```

The important variable here is `MobileHome` which represents the percentage of homes that are mobile homes in each county.

### Question 2
Next, we will create an interactive preview of these data using the `mapview()` function.

```{r p1-q2}
mapview(mobileHome)
```

We can observe that these data are to the extent of the Greater St. Louis Metropolitan Area.

## Part 2: Static Mapping for Digital Use
### Question 3
Now, we will create a static map using `ggplot2`.

Here is the process broken down:

1. Make a call to the `ggplot()` function

2. Add layers with the `geom_sf()` function, specifying the name of your data

3. Create an aesthetic mapping using `MobileHome` as the fill variable

4. Set the fill to NA for the counties, and choose an appropriate color for the outline

5. Add a `scale_fill_` argument. In this example, I used viridis, but you can use RColorBrewer with `scale_fill_distiller()`

6. Add a name to your scale, as well as a title, subtitle and caption.

7. Add the `theme_minimal()` to hide some default `ggplot2` elements

8. Save the plot to a new object, and print the object after.


```{r p2-q3}
# create a ggplot object
ggplot() +
  geom_sf(data = mobileHome, mapping = aes(fill = MobileHome)) +
  geom_sf(data = counties, fill = NA, color = "#A3A3A3") +
  scale_fill_viridis(option = "magma", name = "Percent of\nMobile Homes", direction = -1) +
  labs(
    title = "Mobile Homes by County",
    subtitle = "St. Louis Metropolitan Region",
    caption = "Map by Branson Fox"
  ) +
  theme_minimal() -> ggplot_mbHome

# print the object
ggplot_mbHome
```

### Question 4
Now, we want to save the map we created. To do this, we will use the `ggsave()` function along with `here()` to specify what we would like to save, in what format, and where.

```{r p2-q4}
ggsave(here("results", "ggplot_mbHome.png"), ggplot_mbHome, dpi = 500)
```

## Part 3: Static Mapping for Print Use
### Question 5
We will now create a static map for print use using the `tmap` package.

Here is the process broken down:

1. Use `tm_shape()` to add layers with your data

2. `tm_polygons()` is used to create a thematic fill, specifying your fill variable to `col =` and then choosing your palette and style and adding a legend title.

3. For the `counties` data, we add it with `tm_shape()` and specify the line width and color to `tm_borders()`

4. We add a scale bar with `tm_scale_bar()`

5. We add a title, remove the border and position the legend in `tm_layout()`

6. Assign and print the object

```{r p3-q5}
# create tmap object
tm_shape(mobileHome) +
  tm_polygons(
    col = "MobileHome",
    palette = "viridis",
    style = "jenks",
    title = "Percent of\nMobile Homes"
    ) +
tm_shape(counties) +
  tm_borders(lwd = 3, col = "#666666") +
  tm_scale_bar() +
  tm_layout(
    title = "Mobile Homes",
    frame = FALSE,
    legend.outside = TRUE,
    legend.position = c("left", "bottom")) -> tmap_mbHome
  
# print object
tmap_mbHome
```

### Question 6
Now, we will save the map we created with `tmap_save`. Again, we will use `here()` to specify where the file should be exported.

```{r p3-q6}
tmap_save(tm = tmap_mbHome, filename = here("results", "tmap_mbHome.pdf"), dpi = 500)
```

### Question 7
We will create the same map again, but change our color palette and add a histogram this time. We will use of the RColorBrewer palettes, in this example `PuBu` and add the argument `legend.hist = TRUE` to the `tm_polygons()` function.

```{r p3-q7}
# create tmap object
tm_shape(mobileHome) +
  tm_polygons(
    col = "MobileHome",
    palette = "PuBu",
    style = "jenks",
    title = "Percent of\nMobile Homes",
    legend.hist = TRUE
    ) +
tm_shape(counties) +
  tm_borders(lwd = 3, col = "#666666") +
  tm_scale_bar() +
  tm_layout(
    title = "Mobile Homes",
    frame = FALSE,
    legend.outside = TRUE,
    legend.position = c("left", "bottom")) -> tmap_mbHome_hist
  
# print object
tmap_mbHome_hist
```

### Question 8
We will save the map in the same format as before, but with a different name this time.

```{r p3-q8}
tmap_save(tm = tmap_mbHome_hist, filename = here("results", "tmap_mbHome_hist.pdf"), dpi = 500)
```

